### 2025-09-24 paper/abstract.tex

\begin{abstract}
We introduce \emph{transparent documents}, interactive web-based scholarly articles which allow readers to
explore the relationship to the underlying data by hovering over fragments of text. We present an agent-based
LLM framework for authoring transparent documents, building on recent developments in data provenance for
general-purpose programming languages. Our implementation uses Fluid, an open source functional programming
language with a provenance-tracking runtime, as a target platform. Our tool consists of two LLM agents which
support a human author during the creation of a transparent document. A \SuggestionAgent helps identify
fragments of text which could plausibly be computed from data, including numerical values selected from
records or computed by aggregations like sum and mean, comparatives and superlatives like ``better than'' and
``largest'', trend-adjectives like ``growing'', and other idiomatic quantitative or semi-quantitative phrases.
An \InterpretationAgent, given such a fragment, then attempts to synthesise a suitable Fluid query over the
data which will generate the target string. The resulting expression is spliced into the source code for an
interactive web page, turning the static text fragment into an interactable data-driven element able to reveal
the data that underwrites the natural language claim. We evaluate our approach on a subset of SciGen, an open
source dataset consisting of tables from scientific articles and their corresponding descriptions, which we
extend with hand-generated counterfactual test cases to evaluate how well machine-generated expressions
generalise in the presence of changes to the data. Our results show that gpt4o is often able to synthesise
compound expressions extensionally compatible with our gold solutions.

\mdnote{The abstract feels a little longer than usual. Maybe it could be shortened slightly (for instance, the part on the two agents), though you may well consider every detail necessary. In any case, please ignore this if it’s too late in the process to revise.}
\end{abstract}

>>>>

\begin{abstract}
We introduce \emph{transparent documents}, interactive web-based scholarly articles which allow readers to explore the relationship to the underlying
data by hovering over fragments of text, and present an LLM-based tool for authoring transparent documents, building on recent developments in data
provenance for general-purpose programming languages. As a target platform, our implementation uses Fluid, an open source programming language with a
provenance-tracking runtime. Our agent-based tool supports a human author during the creation of transparent documents, identifying fragments of text
which can be computed from data, such as numerical values selected from records or computed by aggregations like sum and mean, comparatives and
superlatives like ``better than'' and ``largest'', trend-adjectives like ``growing'', and similar quantitative or semi-quantitative phrases, and then
attempts to synthesise a suitable Fluid query over the data which generates the target string. The resulting expression is inserted into the article's
web page, turning the static text fragment into an interactable data-driven element able to reveal the data that underwrites the natural language
claim. We evaluate our approach on a subset of SciGen, an open source dataset consisting of tables from scientific articles and their corresponding
descriptions, which we extend with hand-generated counterfactual test cases to evaluate how well machine-generated expressions generalise. Our results
show that gpt4o is often able to synthesise compound expressions extensionally compatible with our gold solutions.
\end{abstract}

### 2025-09-24 paper/sec/evaluation.tex
rough pass over future work.

@@

\subsection{Counterfactual testing}
\label{subsec:validation-with-counterfactual-data}

>>>>

\cdnote{I've bulked up the RQs qlthough I'm not sure we're actually attempting to answer them}
RQ1. How robust is the system when provided with incomplete or ambiguous information?
In real-world authoring tasks, text fragments may omit key criteria (e.g. \textbf{?}) or contain vague identifiers (e.g. \textbf{?} without further context).
We therefore ask how performance degrades when the specification is underspecified or when multiple interpretations are possible. Here, we distinguish between two forms of ambiguity:
(1) domain-level ambiguity, which reflects genuine uncertainty in the data (e.g. two models achieving identical scores).
(2) specification-level ambiguity, which arises from incomplete or imprecise descriptions in the query (e.g. missing fields that would otherwise resolve the expression uniquely).

RQ2. How sensitive is the system to misleading or adversarial information?
Authors may, intentionally or unintentionally, introduce misleading elements into the text. These include inconsistent identifiers, incorrect numerical values, or contradictory claims.
We ask how the system responds in such cases: does it fail gracefully, does it default to literal reproduction of the misleading fragment, or does it attempt to generate a plausible, but incorrect, expression?

Together, these questions probe the limits of the InterpretationAgent's ability to generate reliable code. By contrasting genuine uncertainty with misleading or ill-specified inputs, we aim to understand not only the model's raw success rate, but also the failure modes most likely to arise in practical use.

@@

\subsection{Counterfactual testing}
\label{sec:validation-with-counterfactual-data}
